---
title: "SAS with Rmarkdown"
author: "Melinda K. Higgins, PhD."
date: "September 5, 2017"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)

# see what the initial knitr engine settings are
knitr::opts_chunk$get()$engine
knitr::opts_chunk$get()$engine.path
knitr::opts_chunk$get()$engine.opts

```

## Resource Link

To get started using `SAS` as your statistical software/data processing "engine" take a look at the following article: [http://www.ssc.wisc.edu/~hemken/SASworkshops/Markdown/SASmarkdown.html](http://www.ssc.wisc.edu/~hemken/SASworkshops/Markdown/SASmarkdown.html).

Also read up on the `SASmarkdown` package [https://cran.r-project.org/web/packages/SASmarkdown/](https://cran.r-project.org/web/packages/SASmarkdown/).

## Setup

To get started you need:

1. Have SAS installed locally on your machine (i.e. you need a licensed copy)
2. you need to know where on your local drive that your SAS executable is located. Mine is located at `C:\Program Files\SASHome\SASFoundation\9.4\sas.exe`.
3. Install the `SASmarkdown` package.
4. Then setup your `knitr` options as follows:

```
saspath <- "C:/Program Files/SASHome/SASFoundation/9.4/sas.exe"
sasopts <- "-nosplash -linesize 75"
knitr::opts_chunk$set(engine="sashtml", engine.path=saspath, 
        engine.opts=sasopts, comment=NA)
```

## Change settings to use HTML output from SAS

NOTE: You will need to install the `SASmarkdown` package first. For this exercise we're using `engine="sashtml"` which works well when `knitting` to HTML. However, this format will not work if you `knit` to PDF or DOC. This leverages the ODS output from SAS.


```{r}
# load the SASmarkdown package
library(SASmarkdown)

# set up the options so that knit knows where you SAS executable is
# set the linesize to be easily readable on letter size paper, portrait
# and set the knir options using opts_chunk$set().
saspath <- "C:/Program Files/SASHome/SASFoundation/9.4/sas.exe"
sasopts <- "-nosplash -linesize 75"
knitr::opts_chunk$set(engine="sashtml", engine.path=saspath, 
        engine.opts=sasopts, comment=NA)

# run these commands to convince yourself that
# within this knitr session the engine changed.
knitr::opts_chunk$get()$engine
knitr::opts_chunk$get()$engine.path
knitr::opts_chunk$get()$engine.opts
```

## If you want to use `SAS` but make PDF or DOC files

If you want to `knit` to PDF or DOC, you should change the engine to `engine="sas"`. This will NOT use ODS output from SAS.

```
saspath <- "C:/Program Files/SASHome/SASFoundation/9.4/sas.exe"
sasopts <- "-nosplash -linesize 75"
knitr::opts_chunk$set(engine="sas", engine.path=saspath, 
        engine.opts=sasopts, comment=NA)
```

## Try some `SAS` code

This code chunk runs te `PROC MEANS` command from `SAS` using the built in dataset `sashelp.class`.

```{r}
proc means data=sashelp.class;
run;
```

## More SAS code 

... and another `SAS` code chunk using the `PROC CORR` commands to see correlations between the variables and also visualize the scatterplot matrix.

```{r}
proc corr data=sashelp.class plots=matrix;
run;
```

## Reset engine to R

While this is great using SAS, if you want to switch back to using R within the same Rmarkdown document, within the same `knitr` session, you'll need to tell `knitr` that you are switching engines.

To get this next chunk to run, you'll need to reset the `knitr` engine within the chunk options directly and then use the command `knitr::opts_chunk$set(engine="R", engine.path=NULL, engine.opts=NULL, comment=NA)` to reset back to `R`.

```{r, engine='R'}
# this chunk has the engine set back to `R`

# run a short bit of r code
x <- 5
x

# check the current engine
knitr::opts_chunk$get()$engine

# reset the engine globally (i.e. so it will work outside of
# this chunk so you don't have to keep typing engine=`R`)
knitr::opts_chunk$set(engine="R",  engine.path=NULL, 
        engine.opts=NULL, comment=NA)

# confirm that this change was applied
knitr::opts_chunk$get()$engine
knitr::opts_chunk$get()$engine.path
knitr::opts_chunk$get()$engine.opts
```

...and here is a new chunk with no engine defined but the engine has defaulted back to `R`.

```{r}
# this chunk didn't list the engine explicitly, so let's
# make sure the global change carried over into this chunk
knitr::opts_chunk$get()$engine

# and another short bit of R code
y <- 3
y
```


